---
title: "Cut (modular) inference"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Cut (modular) inference}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>", eval = FALSE, purl = FALSE)
```

Joint Bayesian inference lets exposed-period outcomes inform the baseline
(counterfactual) model. When you want the baseline learned from control data
*only* — a "cut" or modular posterior — `countsynth` implements the two-stage
scheme

$$p_{\text{cut}}(\phi, \theta \mid Z, Y) \;=\; p(\phi \mid Z)\;
p(\theta \mid Y, \phi),$$

where $\phi$ is the baseline (factors and fixed effects), $\theta$ the
treatment block, $Z$ the control-cell data, and $Y$ the exposed-cell data.

## How it runs

1. **Stage 1** fits the baseline model on control cells (same priors as the
   joint model with the treatment block off).
2. `num_stage1_draws` posterior draws are selected — chain-stratified,
   seeded, without replacement — and become the *cut components*.
3. **Stage 2** runs one small MCMC per component, with that draw's
   `mu_ctrl` frozen as data, so exposed outcomes can never feed back into
   the baseline.
4. Each component's retained draws are thinned to
   `stage2_draws_per_component` with a deterministic even stride (equal
   counts per component keep pooling weights equal), then pooled with
   provenance columns `cut_component`, `stage1_draw`, `stage1_chain`,
   `stage1_iteration`.

```{r}
library(countsynth)

cfg <- countsynth_example_config(
  model = countsynth_model_opts(
    types = list(total = countsynth_type("total", 3)),
    inference_mode = "cut"
  ),
  cut = countsynth_cut_opts(
    num_stage1_draws = 25,
    stage2_draws_per_component = 100,
    stage2_mcmc = list(iter_warmup = 500, iter_sampling = 1000, thin = 1)
  )
)

cfit <- countsynth_cut_fit(countsynth_data(cfg), config = cfg)
summary(cfit)          # stage-1 gate + per-component diagnostics
draws <- countsynth_draws(cfit)
```

## Seeds

All cut seeds derive from `mcmc$seed`: draw selection uses `seed + 2` and
stage 2 uses `seed + 3` (component $i$ samples at `stage2_seed + i`), unless
set explicitly via `countsynth_cut_opts()`. `stage2_mcmc` must not set a seed —
`stage2_seed` is the authority. Selection is deliberately on its own stream
so it is statistically independent of the draws being selected.

## Diagnostics

Convergence is assessed per real MCMC target — stage 1 once, then each
component separately — and never on the pooled draws (which are a mixture,
not one chain). `cfit$manifest` holds the full gate manifest; write it with
`write_convergence_json()`.

## Parallel stage 2

Stage-2 fits are small and run sequentially by default (each already uses
parallel chains). With many components you can run them on a future plan:

```{r}
future::plan(future::multisession, workers = 4)
cfit <- countsynth_cut_fit(countsynth_data(cfg), config = cfg, parallel = "future")
```
