---
title: "Getting started with countsynth"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Getting started with countsynth}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>", eval = FALSE, purl = FALSE)
```

`countsynth` estimates causal treatment effects on panel count data with a
Bayesian hierarchical model: the untreated log-rate surface is a low-rank
nonnegative factorization plus unit and time fixed effects, and treatment
effects enter hierarchically on exposed cells. Sampling runs in Stan via
cmdstanr.

## The model in one paragraph

For groups $k$, units $d$ (e.g. states), and time periods $n$, observed
counts $y_{kdn}$ are Poisson or Negative Binomial with log-mean

$$\mu^{\text{ctrl}}_{kdn} = \log\Big(\sum_{f=1}^R \text{timefac}_{nfk}\,
w_{dkf}\Big) + \alpha_{kd} + \gamma_{kn} + \log(\text{exposure}_{kdn}),$$

where the $w_{dk\cdot}$ are simplex weights (a Dirichlet-uniform prior),
$\alpha$ are hierarchically pooled unit effects, and $\gamma$ are time
effects. Exposed cells get an additional treatment effect
$\mu = \mu^{\text{ctrl}} + \tau$, with $\tau$ decomposed into per-cell,
per-unit, per-group, and unit-by-group components with half-normal scale
priors. The factorization itself is rotation-non-identifiable by design;
identified quantities (the combined $\mu^{\text{ctrl}}$, $\tau$, $\mu$) are
what the package reports.

## Data format

One CSV, one row per (unit, time), wide columns for each outcome and
denominator:

| state | time | exposed_births | births_total | pop_total | ... |
|-------|------|----------------|--------------|-----------|-----|

Columns are mapped by a schema — either explicit outcome definitions or
prefix-based discovery:

```{r}
library(countsynth)

schema <- countsynth_schema(
  unit_col = "state", 
  time_col = "time", 
  treatment_col = 
  "exposed_births",
  outcomes_from_prefixes = countsynth_prefixes(
    outcome_prefix = "births_", denominator_prefix = "pop_",
    include = "total"
  )
)
```

## Fitting the bundled example

```{r}
cfg <- countsynth_example_config()  # bundled US fertility panel, smoke settings
dat <- countsynth_data(cfg)
fit <- countsynth_fit(dat, config = cfg)

summary(fit)              # convergence gate: R-hat, ESS, divergences
parameter_diagnostics(fit)
```

## Choosing the rank without a rank sweep

The rank `R` is the number of latent factors in the untreated surface, and
choosing it up front is awkward: it is a capacity knob with no natural value,
and the honest answer is usually "whatever the panel supports", which is
exactly what you do not know before fitting.

`rank_shrinkage` turns that into something you read off the fit. Each group
gets a stick-breaking profile over the `R` candidate components saying how
popular each one is, and each unit's weights are that profile nudged and
renormalized. Components the panel cannot use are pushed to near-zero weight
in every unit, so setting `R` too high costs a little wall time instead of
distorting the fit. You set no numbers: the DP mass behind the profile and
the unit-level spread both carry hyperpriors and are sampled.

```{r}
cfg_shrunk <- countsynth_example_config(
  model = countsynth_model_opts(
    outcome_distribution = "NB",
    types = list(total = countsynth_type(groups = "total",
                                         ranks_to_test = 8)),
    rank_shrinkage = TRUE          # defaults: Gamma(2, 1) mass, HalfNormal(1)
  )
)
fit_shrunk <- countsynth_fit(countsynth_data(cfg_shrunk), config = cfg_shrunk)

countsynth_eff_rank_summary(fit_shrunk)   # components actually occupied
countsynth_component_weight_plot(fit_shrunk)  # Bayesian scree plot
countsynth_rank_hyper_plot(fit_shrunk)        # hyperparameters vs their priors
```

`eff_rank[k] = 1 / Σ_r group_weight[k, r]²` is an inverse Simpson index: it
reads *m* when `m` components split the weight evenly. On the bundled panel
it lands near 4 out of 8 — comfortably below the truncation, which is the
signal that `R = 8` was generous enough.

Two things to check rather than assume:

- **Confirm the truncation.** Refit at a larger `R`. If `eff_rank` settles in
  the same place, the rank was generous; if it keeps climbing, the truncation
  is still binding and you have not yet gone high enough.
- **Compare the hyperparameters across groups.** Each group draws its own
  `group_weight_mass` and `unit_weight_sd`, so both panels of
  `countsynth_rank_hyper_plot()` show one density per group. A group sitting
  well away from the others is unlike them — typically one whose counts are
  mostly censored, leaving its units scattered and saying little about how
  many components it needs. Both are per group precisely so that such a group
  does not drag the others with it; seeing it stand apart on the figure is
  the signal to treat its effects with more care than the rest.

## Reporting

```{r}
draws <- countsynth_draws(fit)

countsynth_summary_table(draws)          # observed vs expected, one row per group
countsynth_post_treatment_summary(draws) # per-unit excess, ranked
countsynth_expected_vs_observed(draws)   # per-(unit, time) detail

plot(fit, which = "unit_fit", unit = "Texas")
plot(fit, which = "unit_gap", unit = "Texas")
countsynth_interval_plot(draws, estimand = "ratio", method = "mu")
countsynth_ppc_plots(draws)
```

Or run the whole pipeline (all types × ranks, artifacts on disk) at once:

```{r}
countsynth_run(cfg)
```

## Using existing Python configs

Configs written for the Python `bayesian_panel_nmf` package load directly —
`read_countsynth_config()` accepts the same schema and translates the MCMC key
names (`num_warmup` → `iter_warmup`, `target_accept` → `adapt_delta`, …):

```{r}
cfg <- read_countsynth_config("configs/fertility_smoke_test.yaml")
countsynth_run(cfg)
```
