---
title: "Treatment-effect covariates"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Treatment-effect covariates}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>", eval = FALSE, purl = FALSE)
```

By default the treatment effect on an exposed cell is a hierarchy of
group, unit, and group-by-unit random effects plus an i.i.d. cell-level
term — informative about *how large* the effect is, silent about *what
moves it*. Setting a `treatment_effects` formula replaces that hierarchy
with a linear regression surface, so the effect can vary with covariates:
how it evolves over time since a ban took effect, or how it tracks the
change in distance to the nearest clinic.

The formula applies identically to joint inference and to cut stage 2 —
in cut mode the design is built once and shared by every stage-2
component, and coefficient draws are pooled with the same provenance
columns as the rest of the cut posterior.

## The model

Writing $e$ for an exposed cell, the log-rate on treated cells is
$\mu_e = \mu^{\text{ctrl}}_e + \tau_e$ with

$$\tau_e \;=\; x_e^\top\beta \;+\; \sum_j z_{j,e}^\top u_{j,\ell_j(e)}
\;+\; \sigma_{\text{cell}}\,\varepsilon_e ,$$

where $x_e$ is the fixed-effect design row, each $j$ is a random-effect
term with level $\ell_j(e)$, and $u_{j,\ell} \sim
\mathcal{N}(0, \sigma_j^2)$ shrinks that level's coefficients toward the
regression surface $x_e^\top\beta$. The i.i.d. cell term
$\sigma_{\text{cell}}\varepsilon_e$ (a HalfNormal(0.1) scale) is always
present and never written in the formula.

Setting no formula leaves the original model untouched, down to the
random-number stream: the legacy parameters keep their declaration order
and the regression block is declared zero-size, so an existing analysis
reproduces exactly.

## Syntax

`countsynth_te_opts()` takes an lme4-style one-sided formula. Fixed-effect terms
build the regression surface; each `(expr | g)` term lets `expr`'s
coefficients vary by the levels of `g`, shrinking toward that surface.

```{r}
library(countsynth)

# Effect that trends with time since treatment, varying by group.
countsynth_te_opts(formula = ~ 1 + event_time + (1 + event_time | group))

# Effect mediated by the change in distance to the nearest clinic, with a
# group-specific intercept and slope.
countsynth_te_opts(
  formula = ~ 1 + dist_change + (1 + dist_change | group),
  covariates = "clinic_distance.csv"
)

# Subgroup surfaces that share nothing but the prior: a separate intercept
# and time slope per group, with no pooled fixed part.
countsynth_te_opts(formula = ~ 0 + (1 + event_time | group))
```

These variables are always available:

| Variable | Meaning |
| --- | --- |
| `group` | Outcome group (factor) |
| `unit` | Panel unit (factor), levels restricted to treated units |
| `time_idx` | 1-based period index |
| `event_time` | Periods since the unit's first treated period, 0 at onset |

Grouping factors may be `group`, `unit`, an interaction such as
`group:unit`, or any categorical covariate.

## Covariates

`covariates` takes a data frame or CSV path joined onto the exposed cells
by whichever of `unit`, `time`, `group` it contains. A unit-level mediator
is keyed by `unit` alone; something that also moves over time is keyed by
`unit` and `time`.

Every treated unit needs a value. In the bundled fertility example the
fourteen states with a ban are the ones that carry exposed cells:

```{r}
dist <- data.frame(
  unit = c(
    "Alabama", "Arkansas", "Georgia", "Idaho", "Kentucky", "Louisiana",
    "Mississippi", "Missouri", "Oklahoma", "South Dakota", "Tennessee",
    "Texas", "West Virginia", "Wisconsin"
  ),
  dist_change = c(
    227.8, 195.4, 172.6, 310.5, 158.9, 243.7,
    288.1, 141.2, 178.9, 264.3, 166.5,
    210.4, 132.7, 119.8
  )
)

cfg <- countsynth_example_config(
  model = countsynth_model_opts(
    types = list(total = countsynth_type("total", 3)),
    inference_mode = "cut",
    treatment_effects = countsynth_te_opts(
      formula = ~ 1 + event_time + dist_change + (1 + event_time | unit),
      covariates = dist
    )
  )
)
cfit <- countsynth_cut_fit(countsynth_data(cfg), config = cfg)
```

Three things are worth knowing before you build a design. Only exposed
cells enter it, so a unit-level covariate is constant — and therefore
collinear with the intercept — unless several units are treated; a
rank-deficient design is refused rather than sampled. Covariates must be
non-missing on every exposed cell they are used for. And a grouping factor
with only one level among the exposed cells (grouping by `group` when the
type has a single outcome group, say) warns: its coefficients are
confounded with the matching fixed effects, and only the prior tells them
apart.

Continuous covariates are standardized by default, so a fixed-effect
coefficient reads as the effect per standard deviation. Plots map back to
the original scale regardless; set `standardize = FALSE` to sample on the
raw scale.

## Reading the results

Every existing table and figure is unchanged — the treatment effect still
flows through `mu_treated`, so `countsynth_summary_table()`,
`countsynth_interval_plot()`, and the rest report exactly as before. The
regression adds its own layer on top.

```{r}
# Posterior of the regression function over a covariate, with 67% / 95%
# bands, on the covariate's original scale.
countsynth_te_regression_plot(cfit, predictor = "event_time")

# One surface per subgroup: the fixed surface plus each level's deviation.
countsynth_te_regression_plot(cfit, predictor = "event_time", by = "unit")

# As a percent change in the rate rather than an additive log-rate effect.
countsynth_te_regression_plot(cfit, predictor = "dist_change", scale = "percent")

# Coefficients as a forest plot, and as a table.
countsynth_te_coef_plot(cfit, terms = "all")
countsynth_te_coef_table(cfit)

# The numbers behind the curve, and the raw coefficient draws.
countsynth_te_regression_summary(cfit, predictor = "event_time", by = "unit")
countsynth_te_draws(cfit)
```

In `countsynth_te_draws()` and `countsynth_te_coef_table()`, a `(fixed)` row is the
population coefficient and a random-effect row is that level's *deviation*
from the fixed surface. `countsynth_te_coef_plot()` shows the total per-level
coefficient by default; pass `deviation = TRUE` for the deviations alone.

Adding `"te_regression"` to `output$figures` writes the whole set under
`figs/te/` on a `countsynth_run()`, alongside `te_coefficients.csv`.

```{r}
cfg$output <- countsynth_output_opts(figures = c("interval", "te_regression"))
countsynth_run(cfg)
```

## YAML

```yaml
model:
  inference_mode: cut
  treatment_effects:
    formula: ~ 1 + event_time + dist_change + (1 + event_time | unit)
    covariates_file: clinic_distance.csv
    standardize: true
    coef_prior_scale: 1.0
    re_prior_scale: 1.0
```
